\chapter{Installation and Running}\label{chap:appD_running}

This appendix is the practical companion to the conceptual chapters: how to
install \model, stage its input data, and run it on three increasingly large
targets --- a laptop CPU, a single GPU, and a SLURM cluster. Every command below
is taken from the repository (\code{README.md}, \code{requirements.txt}, and the
SLURM scripts under \code{experiments/}) and should be run from the repository
root unless stated otherwise. Because \model{} is research software, treat these
instructions as a recipe that is validated against the pinned dependency stack
(see Appendix~\ref{chap:appB_parameters} for parameter values and
Appendix~\ref{chap:appC_codemap} for the code map).

\section{Software stack and the JAX pin}

\model{} is a \jax{} program. The single most important constraint is that the
JAX version is \emph{pinned}: the multi-level atmosphere is built on the
\dino{} dycore (\code{dinosaur-dycore==1.2.1}), which uses fast-moving private
JAX APIs and breaks on newer JAX releases. The repository therefore pins

\begin{lstlisting}[language=bash]
jax==0.8.1
jaxlib==0.8.1
dinosaur-dycore==1.2.1
\end{lstlisting}

\noindent Do \emph{not} upgrade JAX to chase a newer feature: doing so silently
breaks the \dino{} core. The remaining dependencies are standard scientific
Python --- \code{numpy}, \code{scipy}, \code{xarray}, \code{netCDF4},
\code{matplotlib}, and \code{pooch} (which manages the automatic data cache,
Section~\ref{sec:appD-data}) \citep{harris2020numpy,virtanen2020scipy,hoyer2017xarray,hunter2007matplotlib,uieda2020pooch,dinosaur_software}. The optional \code{cdsapi} is needed only for
ERA5 validation downloads.

\section{Cloning and a CPU install}

The default install gives you a \emph{CPU} build of JAX, which is the right
choice for a laptop, for the test suite, and for short experiments. Clone the
repository, create a virtual environment, and install the pinned requirements:

\begin{lstlisting}[language=bash]
git clone https://github.com/bijanf/ESMB.git && cd ESMB
python -m venv venv && source venv/bin/activate
pip install -r requirements.txt
\end{lstlisting}

\noindent This installs CPU JAX. Verify the install by asking JAX which devices
it can see; on a CPU-only box you should see a \code{CpuDevice}:

\begin{lstlisting}[language=bash]
python -c "import jax; print(jax.devices())"
\end{lstlisting}

\section{The GPU (CUDA) build}\label{sec:appD-gpu}

\code{requirements.txt} deliberately installs CPU JAX so the package is usable
everywhere. For GPU you must additionally install the matching CUDA wheel
\emph{at the same JAX version} --- mismatched versions are the most common cause
of a silent fall-back to CPU or an import error. The robust idiom reads the
installed version and pins the CUDA build to it:

\begin{lstlisting}[language=bash]
pip install "jax[cuda12]==$(python -c 'import jax; print(jax.__version__)')"
\end{lstlisting}

\noindent This pulls the bundled \code{nvidia-*} wheels, so you do not need a
system-wide CUDA toolkit for the libraries themselves (you still need a
compatible NVIDIA driver). Confirm the GPU is visible before launching any long
run:

\begin{lstlisting}[language=bash]
python -c "import jax; print(jax.devices())"   # expect a CudaDevice / gpu
\end{lstlisting}

\noindent On a single T31 GPU the model runs roughly $280$ simulated years per
day; the coupled \dino{}$\leftrightarrow$ocean control runs near $30$
model-years per hour. The GPU jobs enable double precision via
\code{JAX\_ENABLE\_X64=True}.

\section{Staging the input data}\label{sec:appD-data}

The model's input datasets are \emph{not} committed to the repository. On first
use they download automatically through \code{pooch} and cache to
\code{\textasciitilde/.cache/chronos\_esm/} (\code{pooch.os\_cache("chronos\_esm")}).
Two datasets are required for any run, a third only for validation:

\begin{center}
\begin{tabular}{@{}llrl@{}}
\toprule
Dataset & Used for & Size & Source \\
\midrule
WOA18 T+S ($5^\circ$) & ocean IC + validation & $\sim 8$ MB & NOAA NCEI \\
ETOPO1 bathymetry & land mask + orography & $\sim 900$ MB & SOCIB THREDDS \\
ERA5 monthly means & validation only & $\sim$ MBs & Copernicus CDS \\
\bottomrule
\end{tabular}
\end{center}

\noindent WOA18 supplies the ocean initial $\Temp$/$\Salt$ fields and the
validation target; ETOPO1 supplies the variable-depth bathymetry (the ocean land
mask) and the atmospheric orography. ERA5 is needed only when you pass
\code{--era5} to a validation script, and it requires a Copernicus CDS account
configured in \code{\textasciitilde/.cdsapirc}.

You rarely need to stage data by hand for a local CPU run --- the first model
build downloads what it needs. The explicit staging tool,
\codref{experiments/prefetch_data.py}{the data-cache prefetcher}, exists mainly
for clusters (Section~\ref{sec:appD-slurm}) but is useful anywhere:

\begin{lstlisting}[language=bash]
python experiments/prefetch_data.py          # fetch WOA18 + ETOPO1 into the cache
python experiments/prefetch_data.py --era5   # also fetch ERA5 (needs ~/.cdsapirc)
python experiments/prefetch_data.py --build  # + build the model once (full preflight)
python experiments/prefetch_data.py --check  # verify the cache only; exit 1 if missing
\end{lstlisting}

\noindent If your \code{\$HOME} is space-constrained, redirect the cache by
exporting \code{XDG\_CACHE\_HOME} to a larger filesystem \emph{before} both the
prefetch and the run, so they agree on the path.

\section{Running locally on CPU}\label{sec:appD-cpu}

For development and small experiments, run on CPU and force the platform
explicitly so a stray CUDA install cannot change behaviour. The repository idiom
prepends the repo root to \code{PYTHONPATH} (the experiment scripts also do this
internally, but it makes ad-hoc invocations robust) and pins the platform with
\code{JAX\_PLATFORMS=cpu}:

\begin{lstlisting}[language=bash]
PYTHONPATH=$PWD JAX_PLATFORMS=cpu python experiments/run_century_physics.py \
    --years 1 --ocean-ic woa
\end{lstlisting}

\noindent A one-year (or a few-day) run is the natural smoke test: it exercises
the full coupled step, downloads any missing data, and finishes in minutes on a
laptop. For the multi-level coupled control run, the same pattern applies:

\begin{lstlisting}[language=bash]
# a short coupled dinosaur <-> ocean control run on CPU:
PYTHONPATH=$PWD JAX_PLATFORMS=cpu python experiments/run_dino_coupled.py --days 5
# resume an existing run toward a larger total (here, toward day 36500):
PYTHONPATH=$PWD JAX_PLATFORMS=cpu python experiments/run_dino_coupled.py \
    --years 100 --resume 36500
\end{lstlisting}

\noindent The harness \codref{experiments/run_dino_coupled.py}{the coupled
control-run driver} is checkpointing and resumable: \code{--years}/\code{--days}
give the \emph{absolute} total, \code{--resume <day>} restarts from a saved day,
and \code{--ckpt-every-days} sets the checkpoint cadence (yearly by default).
Each checkpoint writes both the ocean state (\code{state\_d<DAY>.nc}) and the
\dino{} modal atmosphere (\code{state\_d<DAY>\_dino.npz}). Resuming restores the
prognostic state (ocean $\Temp$/$\Salt$ and the modal dynamics) exactly;
diagnostic fields --- the ocean streamfunction warm-start, density $\dens$,
vertical velocity $\ww$, and DIC --- are recomputed, so a restart is
state-complete but not bit-reproducible (a divergence of order $0.03$ K/week,
far below any climate signal).

\section{Running on a single GPU}

On a machine with a GPU, install the CUDA build (Section~\ref{sec:appD-gpu})
and simply omit the \code{JAX\_PLATFORMS=cpu} pin so JAX selects the GPU. The
GPU jobs set double precision and the resolution explicitly:

\begin{lstlisting}[language=bash]
export CHRONOS_RESOLUTION=T31
export JAX_PLATFORM_NAME=gpu
export JAX_ENABLE_X64=True
python experiments/run_dino_coupled.py --years 100 --outdir outputs/dino_control
\end{lstlisting}

\noindent Always confirm the GPU is actually in use before committing to a long
run --- a model that silently fell back to CPU will be orders of magnitude
slower:

\begin{lstlisting}[language=bash]
python -c "import jax; print(jax.devices())"   # must list a gpu device
\end{lstlisting}

\section{Submitting to a SLURM cluster}\label{sec:appD-slurm}

Compute nodes on most clusters have \emph{no internet}, so the workflow is:
stage the data once on a login node (which does have internet), then submit the
job, which reads from the shared cache. The full login-node sequence is:

\begin{lstlisting}[language=bash]
git clone https://github.com/bijanf/ESMB.git && cd ESMB
module load anaconda cuda/12.3.1            # site-specific module names
python -m venv venv && source venv/bin/activate
pip install -r requirements.txt
# GPU JAX is REQUIRED on the cluster (the SLURM script aborts without a GPU):
pip install "jax[cuda12]==$(python -c 'import jax; print(jax.__version__)')"

# stage WOA18 + ETOPO1 into ~/.cache/chronos_esm on the LOGIN node:
python experiments/prefetch_data.py         # add --era5 to also stage validation data
\end{lstlisting}

\noindent Then submit. The control-run script,
\codref{experiments/run_dino_control_slurm.sh}{the SLURM control-run wrapper},
forwards all arguments to \code{run\_dino\_coupled.py}:

\begin{lstlisting}[language=bash]
sbatch experiments/run_dino_control_slurm.sh --years 100
tail -f logs/dino_control_*.log
\end{lstlisting}

\subsection{What the SLURM script does for you}

The wrapper encodes several cluster conventions that are worth understanding,
because you will likely need to adapt them to your site. Its resource request is
a single GPU node:

\begin{lstlisting}[language=bash]
#SBATCH --partition=gpu
#SBATCH --qos=gpumedium
#SBATCH --account=poem
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=8
#SBATCH --gres=gpu:1
#SBATCH --mem=64G
#SBATCH --time=23:00:00
\end{lstlisting}

\noindent \textbf{Adjust for your site:} the \code{--account}, \code{--partition},
and \code{--qos} lines (the defaults \code{poem}/\code{gpu}/\code{gpumedium} are
PIK-specific) and the \code{module load} names. The script then:

\begin{enumerate}
\item \textbf{Loads modules and the venv.} It loads \code{python/3.12.3} (must
match the Python the venv was built with) and \code{cuda/12.6.0} (must match the
\code{jax[cuda12]} build), then activates \code{venv}. It also prepends the
bundled \code{nvidia-*} wheel library paths to \code{LD\_LIBRARY\_PATH}.

\item \textbf{Runs a GPU preflight.} A tiny Python snippet asserts that JAX sees
a GPU device and aborts the job immediately if not, so a multi-hour allocation
is never wasted on a CPU fall-back.

\item \textbf{Runs a data-cache guard.} It calls
\code{prefetch\_data.py --check}; if WOA18/ETOPO1 are not staged, the job fails
fast with instructions to run the prefetch on a login node, rather than hanging
on a blocked download.

\item \textbf{Writes output to scratch, never \texttt{\$HOME}.} Output goes to
\code{\$CHRONOS\_OUTDIR} (PIK default \code{/p/tmp/\$USER/dino\_control});
override it by exporting \code{CHRONOS\_OUTDIR} to your own scratch path.

\item \textbf{Auto-resumes across jobs.} If you do not pass \code{--resume}, the
script scans the output directory for the latest \code{state\_d<DAY>.nc}
checkpoint and resumes from it. This is what lets you chain back-to-back $23$-hour
jobs against a single \code{--years} target until the run completes:
\end{enumerate}

\begin{lstlisting}[language=bash]
# first job (fresh): runs until walltime, checkpointing yearly to scratch
sbatch experiments/run_dino_control_slurm.sh --years 100
# subsequent jobs (same command): AUTO-RESUME from the latest checkpoint
sbatch experiments/run_dino_control_slurm.sh --years 100
# explicit resume overrides the auto-detect:
sbatch experiments/run_dino_control_slurm.sh --years 200 --resume 36500
\end{lstlisting}

\noindent If your \code{\$HOME} quota is small, export
\code{XDG\_CACHE\_HOME=/path/on/shared/scratch} \emph{before both} the prefetch
and the \code{sbatch} --- the cache must be the same path, visible from both the
login and compute nodes.

\subsection{The AMOC hosing sweep}

A second SLURM wrapper,
\codref{experiments/run_amoc_hosing_slurm.sh}{the AMOC hosing/hysteresis sweep},
drives the freshwater-hosing experiment behind the bistability and tipping
results of Chapter~\ref{chap:ch17_amoc_tipping}. It uses the same module/venv/GPU
preflight preamble but requests a preemptible QOS and forwards all arguments to
\code{run\_amoc\_hosing.py}, which auto-resumes from per-hosing-level
checkpoints (so a \code{--requeue} after preemption continues rather than
re-spinning up):

\begin{lstlisting}[language=bash]
sbatch --dependency=afterok:<control_jobid> --requeue \
    experiments/run_amoc_hosing_slurm.sh \
    --ckpt /p/tmp/$USER/dino_control_thc/state_d036500 \
    --fmax 1.0 --nsteps 5 --hold-years 15 \
    --outdir /p/tmp/$USER/amoc_hosing
\end{lstlisting}

\noindent The sweep ramps the subpolar freshwater hosing up to \code{--fmax}
(in \Sv) in \code{--nsteps} stages, holding each for \code{--hold-years} so the
overturning equilibrates; the \code{--dependency=afterok} chains it after a
finished control run.

\section{Scoring a run against observations}

Once a run has written checkpoints, score them against WOA18 (ocean
$\Temp$/$\Salt$) and ERA5 (near-surface atmosphere) with the validation
pipeline (Chapter~\ref{chap:ch16_climatology}). It accepts a glob of checkpoints,
which it averages into a climatology:

\begin{lstlisting}[language=bash]
# score saved checkpoints (scorecard + bias maps + zonal means):
python experiments/validate_control.py --states "outputs/dino_control/state_d*.nc" --era5
# or exercise the pipeline on a short in-process run:
python experiments/validate_control.py --demo 200 --era5
# refresh the README dashboard figures from a saved glob:
python experiments/make_readme_figures.py "outputs/dino_control/state_d*.nc" \
    --label "dino control"
\end{lstlisting}

\noindent A perfect model scores bias $0$, correlation $1$, and standard-deviation
ratio $1$; the current scorecards (and their honest limitations) are discussed in
Chapters~\ref{chap:ch16_climatology}--\ref{chap:ch19_tropics_monsoon}.

\section{Running the test suite}

The test suite under \code{tests/} is the fastest way to confirm an install is
healthy and to guard changes. It runs on CPU; use the same platform pin as for
local runs:

\begin{lstlisting}[language=bash]
PYTHONPATH=$PWD JAX_PLATFORMS=cpu python -m pytest
\end{lstlisting}

\noindent The tests cover the ocean and atmosphere components, the coupler and
flux correction, the AMOC and momentum diagnostics, the \dino{} coupling and
time step, the CO$_2$ warming end-to-end path, and --- crucially for a
differentiable model --- the adjoint/gradient checks. Run a single module, for
example the adjoint tests, with the usual pytest selector:

\begin{lstlisting}[language=bash]
PYTHONPATH=$PWD JAX_PLATFORMS=cpu python -m pytest tests/test_adjoint.py -q
\end{lstlisting}

\noindent The pytest configuration in \code{pyproject.toml} fixes the test path
to \code{tests/} and silences deprecation warnings, so a bare \code{pytest}
invocation from the repository root behaves identically. With the test suite
green and a short coupled run completing without NaNs, the install is ready for
the experiments of Part~III.
