\chapter{The Ocean: Subgrid Physics and Mixing}\label{chap:ch05_ocean_mixing}

% Local symbols used only in this chapter (kept minimal; not in the preamble).
\providecommand{\Sslope}{\mathbf{S}}          % isopycnal slope vector (S_x, S_y)
\providecommand{\kgm}{\kappa_{\mathrm{gm}}}    % Gent-McWilliams thickness diffusivity
\providecommand{\kiso}{\kappa_{\mathrm{iso}}} % Redi isopycnal diffusivity
\providecommand{\kh}{\kappa_{h}}              % horizontal (Laplacian) tracer diffusivity
\providecommand{\kv}{\kappa_{z}}             % vertical tracer diffusivity

A model run at the T31 ocean grid of \model{} ($\approx 3.75^\circ$, see
Chapter~\ref{chap:ch04_ocean_equations}) cannot resolve the mesoscale eddy
field: the first baroclinic Rossby radius in the mid-latitude ocean is
$\mathcal{O}(20\text{--}50)\,\mathrm{km}$, an order of magnitude below a grid box.
The eddies are nonetheless dynamically essential---they flatten isopycnals, set
the stratification, and carry a large fraction of the meridional heat transport.
This chapter documents how the unresolved physics is \emph{parameterized}: the
Gent--McWilliams (GM) bolus advection \citep{gentmcwilliams1990}, along-isopycnal
(Redi) mixing, horizontal and vertical diffusion, the smooth convective
adjustment, and the Shapiro/biharmonic filter used purely for numerical hygiene.
Throughout, the governing continuous form precedes the discrete operator that the
code actually integrates, with the source pinned by a \emph{Code} marker. The
canonical reference for the rotated tracer-mixing tensor is \citet{griffies2004};
the GM closure is \citet{gentmcwilliams1990}; the standard textbook treatment is
\citet{vallis2017}.

\begin{remark}
All operators below are written in \jax{} and are differentiable. Where a naive
formulation would introduce a non-smooth $\max$, hard clip, or \code{where} step
(zero gradient outside the bound), the code substitutes a smooth surrogate so that
$\partial(\text{output})/\partial(\text{parameter})$ stays well defined---this is a
deliberate design choice, not an accident (Chapter~\ref{chap:ch03_differentiable}).
\end{remark}

\section{Isopycnal slopes}

Both GM and Redi act \emph{along} surfaces of constant density. The local slope of
those surfaces, relative to the geopotential, is the slope vector
$\Sslope = (S_x, S_y)$ with
\begin{equation}
S_x = -\frac{\partial_x \dens}{\partial_z \dens}, \qquad
S_y = -\frac{\partial_y \dens}{\partial_z \dens},
\label{eq:slope}
\end{equation}
where $\dens$ is the in-situ density and $z$ increases downward, so a statically
stable column has $\partial_z\dens>0$. Each gradient is a centred finite
difference; the horizontal differences use \code{jnp.roll} (periodic in $\lon$),
and the vertical gradient is taken at interfaces and averaged back to cell centres
using the centre-to-centre spacing $\tfrac12(\Delta z_k + \Delta z_{k+1})$.
\codref{chronos\_esm/ocean/mixing.py}{\code{compute\_isopycnal\_slopes}}

The denominator is regularized to keep the slope finite in weakly stratified
columns:
\begin{equation}
\partial_z\dens \;\longmapsto\;
\begin{cases}
\partial_z\dens, & |\partial_z\dens| \ge \varepsilon,\\[2pt]
\operatorname{sign}(\partial_z\dens)\,\varepsilon, & |\partial_z\dens| < \varepsilon,
\end{cases}
\qquad \varepsilon = 10^{-8}\,\mathrm{kg\,m^{-4}}.
\label{eq:slope-floor}
\end{equation}
This is an honest approximation: in a near-neutral or unstably stratified column
the ``isopycnal'' slope is not physically meaningful, and \eqref{eq:slope-floor}
simply prevents a blow-up while convective adjustment (\S\ref{sec:convect}) handles
the genuine instability.

\section{The slope limiter}

Steep isopycnals near the surface mixed layer and at boundary currents produce very
large $|\Sslope|$, which would make the GM and Redi terms numerically explosive.
The standard remedy is to taper the slope \citep{danabasoglu1995,large1997}. \model{} replaces the conventional hard
clip $|\Sslope|\le S_{\max}$ with a \emph{smooth, differentiable} soft clip:
\begin{equation}
\widetilde{S} \;=\; \frac{S}{\sqrt{1 + (S/S_{\max})^2}},
\qquad S_{\max} = 10^{-2},
\label{eq:limiter}
\end{equation}
applied component-wise to $S_x,S_y$. As $S\to0$, $\widetilde S\to S$ (no distortion
of gentle slopes); as $S\to\pm\infty$, $\widetilde S\to \pm S_{\max}$ (bounded). Crucially,
$\mathrm{d}\widetilde S/\mathrm{d}S>0$ everywhere, so the limiter never zeroes the
gradient---a hard clip would, breaking adjoint sensitivity through steep regions.
\codref{chronos\_esm/ocean/mixing.py}{\code{slope\_limiter}}

\section{Gent--McWilliams bolus transport}

The GM parameterization \citep{gentmcwilliams1990} represents the eddy-induced
(``bolus'') transport as an extra advective velocity $\uvec^\star=(u^\star,v^\star,w^\star)$
that adiabatically flattens isopycnals, extracting available potential energy as a
mesoscale eddy field would. With a thickness diffusivity $\kgm$ the eddy streamfunction
is $\boldsymbol{\Psi}=\kgm\,\Sslope$, and the bolus velocity is its curl:
\begin{align}
u^\star &= \partial_z\!\left(\kgm\,\widetilde S_x\right), &
v^\star &= \partial_z\!\left(\kgm\,\widetilde S_y\right), \label{eq:gm-uv}\\
w^\star &= -\left[\partial_x\!\left(\kgm\,\widetilde S_x\right) + \partial_y\!\left(\kgm\,\widetilde S_y\right)\right].
\label{eq:gm-w}
\end{align}
By construction $\diver\uvec^\star=0$, so GM transports tracers without creating or
destroying mass.

In the code, the horizontal bolus fluxes $F_x=\kgm\widetilde S_x$, $F_y=\kgm\widetilde S_y$
are interpolated to vertical interfaces and set to zero at the surface and floor
(no bolus flux through the boundaries); the vertical derivative
\eqref{eq:gm-uv} is then a centred difference over the layer thickness $\Delta z_k$:
\begin{equation}
u^\star_k = \frac{F_x^{\,k+1/2} - F_x^{\,k-1/2}}{\Delta z_k}, \qquad
F_x^{\,k+1/2} = \tfrac12\big(F_x^{k}+F_x^{k+1}\big),
\label{eq:gm-disc}
\end{equation}
and likewise for $v^\star$.
\codref{chronos\_esm/ocean/mixing.py}{\code{compute\_gm\_bolus\_velocity}}

\begin{remark}[Honest limitation: $w^\star$ is not formed here]
The function returns $w^\star\equiv 0$. The vertical bolus velocity is \emph{not}
diagnosed from \eqref{eq:gm-w} in \code{mixing.py}; instead the bolus velocity is
\emph{added to the resolved flow}---$\uvec_{\mathrm{eff}}=\uvec+\uvec^\star$ with
$u^\star,v^\star$ from \eqref{eq:gm-disc}---and $w$ is recovered from continuity of
that combined field, so the three-dimensional transport that advects tracers is
nondivergent by construction.
\codref{chronos\_esm/ocean/veros\_driver.py}{$\uvec_{\mathrm{eff}}=\uvec+\uvec^\star$; $w$ from $\diver\uvec_{\mathrm{eff}}=0$}
This is consistent but means GM enters the model as a velocity, not as a skew-flux on
each tracer separately---adequate for a coarse climate run, but not the full
Griffies skew-diffusion form.
\end{remark}

The diffusivity $\kgm$ defaults to $1000\,\mathrm{m^2\,s^{-1}}$ in physics mode
(versus $3000$ in the legacy ``tank'' configuration; see the parameter table in
Appendix~\ref{chap:appB_parameters} and the discussion in CLAUDE-context Phase~6).
It is masked to zero in the five polar rows at each end of the grid,
$\kgm^{\mathrm{eff}} = \kgm\,\chi_{\mathrm{int}}$, where $\chi_{\mathrm{int}}=0$ for
$j<5$ or $j\ge n_y-5$ and $1$ otherwise, because the slope estimate degrades where
the grid converges.
\codref{chronos\_esm/ocean/veros\_driver.py}{\code{kappa\_gm\_eff = kappa\_gm * interior\_mask\_3d}}

\section{Along-isopycnal (Redi) mixing tensor}

Diffusion of heat and salt in the ocean interior is overwhelmingly \emph{along}
isopycnals, not across them; diapycnal mixing is weaker by $\mathcal{O}(10^7)$.
The Redi tensor \citep{redi1982} rotates an isotropic diffusivity $\kiso$ into the local isopycnal
plane. In the small-slope approximation \citep{griffies2004},
\begin{equation}
\mathbf{K} \;=\; \frac{\kiso}{1+S^2}
\begin{pmatrix}
1 & 0 & \widetilde S_x\\
0 & 1 & \widetilde S_y\\
\widetilde S_x & \widetilde S_y & S^2
\end{pmatrix},
\qquad S^2 = \widetilde S_x^{\,2} + \widetilde S_y^{\,2},
\label{eq:redi}
\end{equation}
so the diffusive tracer flux is $\mathbf{F} = -\mathbf{K}\,\grad C$ for a tracer $C$.
The diagonal $(1,1)$ block gives near-horizontal diffusion; the off-diagonal
$\widetilde S_x,\widetilde S_y$ couple horizontal gradients to the vertical flux (so
mixing follows the tilted surface); and the $(3,3)$ entry $\propto S^2$ is the small
residual along-slope vertical component. The prefactor $1/(1+S^2)$ is the small-slope
normalization. The tensor is assembled exactly as \eqref{eq:redi} (symmetric,
built from the limited slopes).
\codref{chronos\_esm/ocean/mixing.py}{\code{rotate\_tensor}}

\section{Horizontal and vertical diffusion as integrated}\label{sec:hdiff}

While \code{rotate\_tensor} provides the full Redi operator, the tracer time-stepper
in the active driver uses a \emph{simpler} along-axis diffusion: a horizontal
Laplacian with coefficient $\kh$ plus a vertical diffusion with the convective
$\kv$ of \S\ref{sec:convect}. This is an honest simplification---it is robust on the
coarse grid and the dominant interior balance there is set by GM, not Redi---and it
is what the model integrates today.

The horizontal operator combines upwind advection by $\uvec_{\mathrm{eff}}$ with a
five-point Laplacian,
\begin{equation}
\mathcal{H}(C) = -\big(u_{\mathrm{eff}}\,\partial_x C + v_{\mathrm{eff}}\,\partial_y C\big)
+ \kh\!\left(\frac{C_E - 2C + C_W}{\Delta x^2} + \frac{C_N - 2C + C_S}{\Delta y^2}\right),
\label{eq:hdiff}
\end{equation}
where face neighbours $C_{E,W,N,S}$ revert to the centre value across any \emph{closed}
(land/floor) face, enforcing a no-flux boundary with no exchange into dry cells.
\codref{chronos\_esm/ocean/veros\_driver.py}{\code{\_horiz}, \code{\_nbrs}}
The effective coefficient is $\kh=500\,\mathrm{m^2\,s^{-1}}$ in the interior but is
forced up to $2\times10^4\,\mathrm{m^2\,s^{-1}}$ in the polar rows,
$\kh^{\mathrm{eff}} = 2\!\times\!10^4\,\chi_{\mathrm{pole}} + \kh\,\chi_{\mathrm{int}}$,
to damp grid-scale noise where the meridian convergence is worst.
\codref{chronos\_esm/ocean/veros\_driver.py}{\code{diff\_coef = 20000.*pole\_mask + kappa\_h*interior\_mask}}

Vertical diffusion is flux-form across interfaces,
\begin{equation}
\mathcal{V}(C)_k = \frac{1}{\Delta z_k}\Big(\mathcal{F}^{k-1/2} - \mathcal{F}^{k+1/2}\Big),
\qquad
\mathcal{F}^{k+1/2} = -\,\kv^{\,k+1/2}\,\frac{C_{k+1}-C_k}{\tfrac12(\Delta z_k+\Delta z_{k+1})},
\label{eq:vdiff}
\end{equation}
with $\mathcal{F}=0$ at the surface and floor interfaces. The whole tracer update
($\mathcal{H}+\mathcal{V}$ plus vertical advection and surface fluxes) is advanced by
RK4 with a $\Delta t=900\,\mathrm{s}$ ocean step.
\codref{chronos\_esm/ocean/veros\_driver.py}{\code{\_vdiff}; RK4 \code{tendencies}}

\section{Smooth convective adjustment}\label{sec:convect}

When a column becomes statically unstable ($\partial_z\dens<0$, denser water over
lighter) the ocean overturns and mixes rapidly. The classical implementation is a
hard switch---set $\kv$ to a huge value wherever $\Delta\dens<0$ \citep{rahmstorf1993}. \model{} rejects
this for two reasons documented in the source: (i) a $0/1$ switch fires patchily
column-by-column and produces grid-scale SST noise that grows toward the poles, and
(ii) a large $\kv$ at the thin surface layer violates the explicit-diffusion
stability limit $\kv\Delta t/\Delta z^2 \le \mathcal{O}(1)$.

Instead the diffusivity \emph{ramps smoothly} with the degree of instability
$I = \max(-\Delta\dens,\,0)$ at each interface:
\begin{equation}
\kv = \kappa_{\mathrm{bg}} + (\kappa_{\mathrm{cv}}-\kappa_{\mathrm{bg}})
\Big(1 - e^{-I/I_0}\Big),
\label{eq:convect-ramp}
\end{equation}
with background $\kappa_{\mathrm{bg}}=10^{-5}\,\mathrm{m^2\,s^{-1}}$, convective
ceiling $\kappa_{\mathrm{cv}}=10\,\mathrm{m^2\,s^{-1}}$, and instability scale
$I_0=0.02\,\mathrm{kg\,m^{-3}}$. The factor $1-e^{-I/I_0}$ runs continuously from $0$
(stable) to $1$ (strongly unstable), so neighbouring columns receive
continuously-varying mixing and the operator is smooth (AD-friendly).

Each interface is then capped at its local explicit-stability limit, using the
thinner of the two adjacent layers,
\begin{equation}
\kv \;\longmapsto\; \min\!\left(\kv,\ 0.45\,\frac{\big(\min(\Delta z_k,\Delta z_{k+1})\big)^2}{\Delta t}\right),
\label{eq:convect-cap}
\end{equation}
so thin surface layers stay numerically stable while thick abyssal layers can still
mix strongly---genuine deep convection is preserved.
\codref{chronos\_esm/ocean/mixing.py}{\code{compute\_vertical\_diffusivity}}

\begin{example}[Why the cap matters]
With $\Delta t=900\,\mathrm{s}$ and a surface layer $\Delta z_0 = 10\,\mathrm{m}$,
the cap is $0.45\cdot 10^2/900 = 0.05\,\mathrm{m^2\,s^{-1}}$---two hundred times
\emph{smaller} than $\kappa_{\mathrm{cv}}=10$. Without \eqref{eq:convect-cap} the
explicit scheme would amplify a surface perturbation each step. For a deep layer
$\Delta z=500\,\mathrm{m}$ the cap is $0.45\cdot 500^2/900\approx 125\,\mathrm{m^2\,s^{-1}}\gg\kappa_{\mathrm{cv}}$,
so deep convection is untouched. The single line \eqref{eq:convect-cap} thus reconciles
surface stability with deep mixing.
\end{example}

\section{The Shapiro / biharmonic filter}\label{sec:shapiro}

Centred advection on a C-grid admits a $2\Delta x$ ``checkerboard'' mode that carries
no physical signal but is invisible to Laplacian diffusion at large scales. \model{}
removes it with a scale-selective \emph{biharmonic} (Shapiro) filter \citep{shapiro1970} applied to the
new $\Temp$ and $\Salt$ after each RK4 update:
\begin{equation}
C \;\longmapsto\; C - \frac{\sigma}{16}\,\lap\!\big(\lap C\big),
\label{eq:shapiro}
\end{equation}
where $\lap$ is the five-point Laplacian and $\sigma$ is \code{shapiro\_strength}.
The eigenvalue of $\lap$ for a pure $2\Delta x$ mode is $-4$ per direction, so the
$2\Delta x$ checkerboard has $\lap\lap$-eigenvalue $\approx 16$ and the choice $\sigma=0.25$
\emph{exactly} annihilates it in one pass, while resolved gradients (small
$\lap\lap$) are nearly untouched.
\codref{chronos\_esm/ocean/veros\_driver.py}{\code{compute\_shapiro\_filter}}

\begin{remark}[An honest correction to the legacy default]
An earlier $\nabla^2$ form ($C+\tfrac{\sigma}{4}\lap C$) applied every step was a
strong low-pass smoother that erased real SST/SSS structure---the WOA initial
gradients collapsed toward uniform within days. The biharmonic form
\eqref{eq:shapiro} is far gentler and is \textbf{off by default}
($\sigma=0$): the primary grid-noise control is the physical Laplacian diffusion
($\kh$ on tracers, $A_h$ on momentum, with $A_h=2\times10^5\,\mathrm{m^2\,s^{-1}}$).
The biharmonic filter is a light $2\Delta x$ cleanup to be switched on only if
checkerboarding reappears. Biharmonic momentum mixing ($A_b$, \code{kappa\_bi}) is
likewise available and defaults to zero.
\end{remark}

\section{Parameter summary}

Table~\ref{tab:mix-params} collects the diffusivities and their roles. Physics-mode
values (the realistic configuration) are shown; the legacy ``tank'' values appear in
Appendix~\ref{chap:appB_parameters}.

\begin{table}[ht]
\centering
\caption{Mixing and filter parameters (physics mode).}
\label{tab:mix-params}
\begin{tabular}{lll}
\toprule
Symbol & Value & Role \\
\midrule
$\kgm$ & $1000\ \mathrm{m^2\,s^{-1}}$ & GM thickness (bolus) diffusivity \\
$\kiso$ & set by caller & Redi isopycnal diffusivity (tensor form) \\
$\kh$ & $500\ \mathrm{m^2\,s^{-1}}$ & horizontal tracer Laplacian (interior) \\
$\kh^{\mathrm{pole}}$ & $2\times10^4\ \mathrm{m^2\,s^{-1}}$ & polar-row tracer diffusion \\
$\kappa_{\mathrm{bg}}$ & $10^{-5}\ \mathrm{m^2\,s^{-1}}$ & background vertical diffusivity \\
$\kappa_{\mathrm{cv}}$ & $10\ \mathrm{m^2\,s^{-1}}$ & convective ceiling \\
$I_0$ & $0.02\ \mathrm{kg\,m^{-3}}$ & convective ramp scale \\
$A_h$ & $2\times10^5\ \mathrm{m^2\,s^{-1}}$ & horizontal momentum viscosity \\
$\sigma$ & $0$ (default) & biharmonic (Shapiro) filter strength \\
$S_{\max}$ & $10^{-2}$ & slope-limiter ceiling \\
\bottomrule
\end{tabular}
\end{table}

The eddy heat transport that GM supplies is the quantity tied most directly to the
overturning (Chapter~\ref{chap:ch08_ocean_overturning}) and to AMOC stability
(Chapter~\ref{chap:ch17_amoc_tipping}); because every operator here is
differentiable, $\partial(\text{AMOC})/\partial\kgm$ is computable by the adjoint,
which the calibration applications of Chapter~\ref{chap:ch15_differentiable_applications}
exploit.

\section*{Exercises}

\begin{exercise}[Limiter versus hard clip]
Implement both the smooth limiter \eqref{eq:limiter} and a hard clip
$\widetilde S=\operatorname{clip}(S,-S_{\max},S_{\max})$ in a small \jax{} script and use
\code{jax.grad} to plot $\mathrm{d}\widetilde S/\mathrm{d}S$ for $S\in[-3S_{\max},3S_{\max}]$.
Confirm that the hard clip has zero gradient for $|S|>S_{\max}$ while
\eqref{eq:limiter} stays strictly positive. Explain, in terms of the adjoint, why a
4D-Var sensitivity through a steep western-boundary current would be lost with the hard clip.
\end{exercise}

\begin{exercise}[GM flattens isopycnals]
Initialize a two-layer channel with tilted isopycnals (constant $S_x$) and integrate
the ocean with $\kgm\in\{0,500,2000\}\,\mathrm{m^2\,s^{-1}}$, all other terms off.
Measure the e-folding time of the slope. Verify it scales as
$\tau\sim L^2/\kgm$ for a slope of horizontal scale $L$, and discuss why
$\kgm=3000$ (tank mode) overflattens the stratification relative to $\kgm=1000$.
\end{exercise}

\begin{exercise}[Convective cap by layer thickness]
Using \eqref{eq:convect-cap} with $\Delta t = 900\,\mathrm{s}$, tabulate the capped
$\kv$ for layer thicknesses $\Delta z\in\{5,10,50,200,500\}\,\mathrm{m}$ and the ramp
value of \eqref{eq:convect-ramp} for $I\in\{0,0.01,0.05,0.2\}\,\mathrm{kg\,m^{-3}}$.
Identify the thickness at which the convective ceiling $\kappa_{\mathrm{cv}}=10$ stops
being the binding constraint. Why does this design let deep convection coexist with a
stable thin surface layer?
\end{exercise}

\begin{exercise}[Is the filter eating the signal?]
Run the ocean for 30 days from a WOA initial state with the biharmonic filter at
$\sigma\in\{0,0.25,1.0\}$ and compute the global SST pattern correlation against the
WOA field. Show that $\sigma=0.25$ removes a $2\Delta x$ checkerboard you inject by hand
while leaving the large-scale pattern correlation essentially unchanged, and that
large $\sigma$ degrades it---motivating the default $\sigma=0$ and reliance on physical
$\kh$.
\end{exercise}
