\chapter{The Atmosphere: Single-Level Spectral Model}\label{chap:ch10_atmosphere_singlelevel}

% --- local symbols (minimal, chapter-scoped) -----------------------
\providecommand{\lnps}{\ensuremath{\pi}}   % log surface pressure, pi = ln p_s
\providecommand{\spechum}{q}               % specific humidity
\providecommand{\divg}{\delta}             % horizontal divergence

This chapter documents \model's \emph{legacy} atmosphere: a single horizontal
level (\code{ATMOS\_GRID.nz}\,$=1$), shallow-water-like spectral model integrated
with forward Euler. It is the atmosphere that produced the validation dashboard in
the README, and it remains the only atmosphere wired into the coupled time step
(\code{main.py:step\_coupled}). Its successor --- a multi-level spectral dynamical
core built on the \dino{} dycore --- is the subject of
Chapter~\ref{chap:ch09_atmosphere_multilevel}; that core fixes the structural
limitations we are honest about here (no baroclinic eddies, weak Intertropical
Convergence Zone). The single-level model is nonetheless worth understanding in
detail: it is small, fully differentiable, and its design choices are a clinic in
what one \emph{can} and \emph{cannot} ask of a barotropic atmosphere.

We follow the textbook structure: continuous governing equation first, then the
discrete form the code actually integrates, including every clamp, filter and
limiter, with a pointer to the source. The reference for the underlying dynamics
is \citet{vallis2017}.

\section{State variables and the prognostic--diagnostic split}

The atmospheric state (\codref{chronos\_esm/atmos/dynamics.py}{class
\code{AtmosState}}) carries six \emph{prognostic} fields, each a single
horizontal slab on the $\mathrm{nlat}\times\mathrm{nlon}$ lat--lon grid:
relative vorticity $\vort$, horizontal divergence $\divg$, temperature $T$,
log surface pressure $\lnps=\ln p_s$, specific humidity $\spechum$, and a carbon
dioxide concentration field (ppm). Velocity is \emph{not} prognostic; instead the
model carries $\vort$ and $\divg$ and reconstructs the wind $\uvec=(\uu,\vv)$ each
step via a Helmholtz (vorticity--divergence) decomposition. The diagnostic fields
$\uu,\vv$, the streamfunction $\psib$, the velocity potential $\chi$, and the
static surface geopotential $\phi_s=\grav z_s$ round out the state.

This is the classic vorticity--divergence formulation of a single-layer atmosphere
\citep[Ch.~5]{vallis2017}. Working in $(\vort,\divg)$ rather than $(\uu,\vv)$ makes
the rotational and divergent parts of the flow cleanly separable and lets us solve
two scalar Poisson problems instead of a coupled vector system.

\section{Wind recovery: the Helmholtz decomposition}

By the Helmholtz theorem any horizontal wind field splits into a non-divergent part
(from a streamfunction $\psib$) and an irrotational part (from a velocity
potential $\chi$):
\begin{equation}
  \uvec = \kvec\times\grad\psib + \grad\chi,
  \qquad
  \uu = -\pd{\psib}{y} + \pd{\chi}{x},
  \qquad
  \vv = \pd{\psib}{x} + \pd{\chi}{y}.
  \label{eq:helmholtz}
\end{equation}
The potentials are recovered from the prognostic fields by inverting two Poisson
equations,
\begin{equation}
  \lap\psib = \vort, \qquad \lap\chi = \divg.
  \label{eq:poisson}
\end{equation}
The code does exactly this. Before inversion both $\vort$ and $\divg$ are passed
through a polar filter (below); the inversions in
Eq.~\eqref{eq:poisson} are solved \emph{directly} by an FFT in longitude combined
with a tridiagonal (Thomas) solve in latitude, not iteratively.

\codref{chronos\_esm/atmos/spectral.py}{\code{inverse\_laplacian} --- rFFT in
$x$, batched tridiagonal solve in $y$ for each zonal wavenumber}

\codref{chronos\_esm/atmos/dynamics.py}{\code{step\_atmos} --- \code{psi =
inverse\_laplacian(vort)}, \code{chi = inverse\_laplacian(div)}, then
Eq.~\eqref{eq:helmholtz}}

\begin{remark}
The longitudinal transform is exact (a real FFT), but the meridional direction
uses fourth-order centred finite differences for gradients and second-order
differences for the Laplacian. \model{} is therefore a \emph{spectral--finite-
difference hybrid}, not a true spherical-harmonic model: there is no Legendre
transform and no triangular truncation. The module docstring is candid that a full
pure-\jax{} spherical-harmonic transform ``is 1000+ lines'' and was deliberately
avoided here. This is the single biggest difference from the multi-level
core of Chapter~\ref{chap:ch09_atmosphere_multilevel}.
\end{remark}

After the inversion the diagnostic winds from Eq.~\eqref{eq:helmholtz} are clipped
to $\pm 80\ \mathrm{m\,s^{-1}}$ and polar-filtered. The clip is a safety net, not a
physical cap: with the wind relaxed toward a $\sim 10\ \mathrm{m\,s^{-1}}$
climatological jet (Section~\ref{sec:windrelax}) the field should never approach
$80\ \mathrm{m\,s^{-1}}$, so hitting the clip flags a developing instability rather
than silently saturating an unphysical jet.

\section{The governing tendencies}

On a single layer the prognostic equations are the shallow-water vorticity and
divergence equations augmented with thermodynamic and moisture conservation. The
continuous forms, with $\eta=\vort+\Cor$ the absolute vorticity and
$\Cor=2\Om\sin\lat$ the Coriolis parameter, are
\begin{align}
  \pd{\vort}{t} &= -\uvec\!\cdot\!\grad\eta - \eta\,\divg + \curl\mathbf{F},
    \label{eq:vort} \\[2pt]
  \pd{\divg}{t} &= -\diver\!\big(R_d\,T\,\grad\lnps\big) - \lap\phi_s
                   + \Cor\,\vort + \diver\mathbf{F},
    \label{eq:div} \\[2pt]
  \pd{T}{t}    &= -\uvec\!\cdot\!\grad T + Q_T,
    \label{eq:temp} \\[2pt]
  \pd{\lnps}{t} &= -\uvec\!\cdot\!\grad\lnps - \divg,
    \label{eq:lnps} \\[2pt]
  \pd{\spechum}{t} &= -\uvec\!\cdot\!\grad\spechum + Q_q,
    \qquad
  \pd{C}{t} = -\uvec\!\cdot\!\grad C + Q_C,
    \label{eq:tracers}
\end{align}
where $R_d=287\ \mathrm{J\,kg^{-1}\,K^{-1}}$ is the gas constant for dry air,
$\mathbf{F}$ is the momentum forcing (friction/relaxation, Section~\ref{sec:windrelax}),
$Q_T,Q_q,Q_C$ are the thermodynamic, moisture and carbon sources, and $C$ is the
CO$_2$ field. Equation~\eqref{eq:lnps} is the layer continuity equation: surface
pressure tendency is set by mass convergence $-\divg$ plus advection of $\lnps$.

A few honest remarks on what these are. The pressure-gradient term in
Eq.~\eqref{eq:div} is written in code as
$\diver(R_d T\,\grad\lnps) = R_d\big(\grad T\!\cdot\!\grad\lnps + T\,\lap\lnps\big)$
plus the orographic forcing $-\lap\phi_s$; this is the single-level analogue of the
geopotential gradient driving the divergent flow. Because there is only one level,
there is \emph{no} vertical advection, \emph{no} vertical shear, and hence no
thermal-wind balance --- a fact that drives the central design compromise of
Section~\ref{sec:windrelax}.

\codref{chronos\_esm/atmos/dynamics.py}{\code{step\_atmos}: \code{dzeta\_dt},
\code{ddiv\_dt}, \code{dt\_dt}, \code{dlnps\_dt}, \code{dq\_dt}, \code{dco2\_dt}}

The advection operator is implemented in flux-free advective form,
$\uvec\!\cdot\!\grad f = \uu\,\partial_x f + \vv\,\partial_y f$, with gradients from
the spectral--FD routine:
\begin{lstlisting}[caption={Advection and the vorticity tendency (abbreviated).}]
def advect(f):
    df_dx, df_dy = spectral.compute_gradients(f, dx, dy)
    return -(u * df_dx + v * df_dy)

eta = state.vorticity + f_cor              # absolute vorticity
dzeta_dt = advect(eta) - eta * state.divergence + curl_forcing
\end{lstlisting}

\section{Forward-Euler time stepping with hyperdiffusion}

The model marches every field with a single forward (explicit) Euler step at
$\Delta t = \code{DT\_ATMOS}$, adding scale-selective $\nabla^4$ hyperdiffusion and
a polar-only $\nabla^2$ term. For a generic field $f$ with tendency $G_f$ from
Eqs.~\eqref{eq:vort}--\eqref{eq:tracers},
\begin{equation}
  f^{\,n+1} = f^{\,n} + \Delta t\Big(G_f - \nu_4^{(f)}\,\nabla^4 f^{\,n}
              + \nu_{\mathrm{pol}}\,\nabla^2 f^{\,n}\Big).
  \label{eq:euler}
\end{equation}
The $\nabla^4$ (bi-Laplacian) operator is the Laplacian applied twice
(\codref{chronos\_esm/atmos/spectral.py}{\code{compute\_bilaplacian}}); in spectral
space it damps a mode of wavenumber $k$ as $k^4$, so it bites hard at the grid
scale ($2\Delta x$ noise) and is essentially invisible to the large-scale
circulation. The coefficients are deliberately \emph{strong} --- the so-called
``tank'' values
$\nu_4^{\vort}=3\times10^{14}$, $\nu_4^{\divg}=6\times10^{14}$,
$\nu_4^{T}=2\times10^{13}$, with weaker values for $\spechum$, $C$ and $\lnps$
(all in SI units of $\mathrm{m^4\,s^{-1}}$).

\begin{remark}[The relaxation/diffusion decoupling]
Why keep diffusion strong when one usually weakens it to ``let the jet develop''?
Because in this model the jet is \emph{not} developed by the dynamics --- it is
\emph{set} by the momentum relaxation of Section~\ref{sec:windrelax}. Since
$\nabla^4$ leaves the large scale untouched, strong hyperdiffusion does not fight
the relaxed jet; it only kills small-scale noise and buys stability. A
four-times-weaker setting was found to let a slow eddy instability grow over
$\sim 25$ days until $|\uu|$ hit the clamp. This decoupling ---
\emph{relaxation $\to$ realism, hyperdiffusion $\to$ stability} --- is the key
organising idea of the single-level model. (Contrast the multi-level core, where
the eddies \emph{are} the physics and diffusion must be weak enough to let them
grow.)
\end{remark}

\paragraph{Implicit divergence damping.}
Linear divergence damping is the standard device for suppressing spurious
gravity-wave and external-mode noise. The naive explicit form
$-r\,\divg$ with $r=5\times10^{-2}\ \mathrm{s^{-1}}$ gives
$\Delta t\,r = 1.5 > 1$, so forward Euler \emph{overshoots} and flips the sign of
$\divg$ every step --- a $2\Delta t$ oscillation that pumps grid-scale checkerboard
noise into $\lnps$. The fix is to apply the damping \emph{implicitly}, which is
unconditionally stable for any rate:
\begin{equation}
  \divg^{\,n+1} \;\leftarrow\; \frac{\divg^{\,n+1}}{1 + \Delta t\,r_{\mathrm{div}}},
  \qquad r_{\mathrm{div}} = 10^{-2}\ \mathrm{s^{-1}}.
\end{equation}
This absorbs divergent-mode noise while preserving the large-scale divergent
(Hadley/Walker) circulation that organises precipitation.

\section{The polar \texorpdfstring{$\Delta x$}{dx}-floor: a stability fix, not a physics choice}
\label{sec:polardx}

A lat--lon grid has a fatal pathology at the poles: the zonal grid spacing
\begin{equation}
  \Delta x(\lat) = \frac{2\pi a\cos\lat}{\mathrm{nlon}}
\end{equation}
collapses to $\sim 4\ \mathrm{m}$ on the pole-most rows (where $\cos\lat$ is floored
to $10^{-5}$ to avoid division by zero). Any zonal structure then makes
$\partial_x$ and $\partial_x^2$ explode: the $\diver(R_d T\grad\lnps)$ pressure
term and $\uvec\!\cdot\!\grad\lnps$ advection detonate \emph{at} the poles within a
few steps. The original symptom was diagnosed precisely: surface pressure swung to
$100$--$1200$ hPa at $\lat=\pm 90^\circ$ over flat topography, with
$\sigma(p_s)\sim 150$ hPa against an observed $\sim 12$ hPa.

The fix is to floor the \emph{dynamical} $\Delta x$ at its $80^\circ$ value,
\begin{equation}
  \Delta x_{\mathrm{dyn}}(\lat) = \frac{2\pi a}{\mathrm{nlon}}\,
     \max\!\big(\cos\lat,\;\cos 80^\circ\big),
  \label{eq:dxfloor}
\end{equation}
so the handful of pole-most rows integrate with the $\sim 80^\circ$ metric instead
of the collapsing one. A separate, identical floor $\Delta x_{\mathrm{diff}}$ is
applied to the diffusion operators, whose $1/\Delta x^4$ scaling would otherwise
exceed the explicit stability limit by $\sim 14$ orders of magnitude at the poles.
The \emph{accurate} $\cos\lat$ is retained for area weighting and the Coriolis
parameter. This is unapologetically a numerical band-aid for the grid choice; it is
local to a few rows and does not touch the physics elsewhere.

\codref{chronos\_esm/atmos/dynamics.py}{\code{step\_atmos}: \code{cos\_lat\_dyn},
\code{cos\_lat\_diff}, and \code{polar\_filter} usage}

The polar \emph{filter} (\codref{chronos\_esm/atmos/spectral.py}{\code{polar\_filter}})
is the complementary device: it applies a smooth Gaussian taper to zonal modes
above the CFL cutoff $m_{\max}(\lat)=\tfrac{\mathrm{nlon}}{2}\cos\lat$, always
preserving at least the zonal mean ($m=0$). It \emph{damps} rather than removes
high modes --- an earlier hard cutoff killed all polar circulation every step.

\section{Wind relaxation: an eddy-momentum-flux parameterization}
\label{sec:windrelax}

Here is the central honesty of the chapter. A single-level (barotropic) atmosphere
has no vertical shear, hence no thermal wind and no baroclinic instability. A
meridional temperature gradient therefore drives \emph{no} sustained pressure
gradient: the winds geostrophically adjust toward rest. In diagnosis,
$|\uu|$ decayed from $10$ to $<1\ \mathrm{m\,s^{-1}}$ and the correlation against
ERA5 surface wind \citep{hersbach2020era5} fell to $\sim 0$. The eddy-momentum-flux convergence that builds
the real mid-latitude surface westerlies, and the Hadley overturning that drives
the trades, both live in vertical structure this model does not resolve.

\model{} parameterizes their \emph{net} effect as a Rayleigh relaxation of the wind
toward an observed zonal-mean surface profile, rather than toward zero:
\begin{equation}
  \mathbf{F} = -\frac{\uu - u_{\mathrm{tgt}}(\lat)}{\tau_f}\,\hat{\mathbf{x}}
               -\frac{\vv}{\tau_f}\,\hat{\mathbf{y}},
  \qquad \tau_f = 2\ \text{days}.
  \label{eq:drag}
\end{equation}
The target is a three-Gaussian fit to the observed surface wind: mid-latitude
westerlies, tropical trade easterlies, and weak polar easterlies, with $\lat$ in
degrees,
\begin{equation}
  u_{\mathrm{tgt}}(\lat) =
     9.0\,e^{-\big(\frac{|\lat|-48}{18}\big)^2}
   - 5.5\,e^{-\big(\frac{|\lat|-13}{11}\big)^2}
   - 2.0\,e^{-\big(\frac{|\lat|-80}{10}\big)^2}.
  \label{eq:utarget}
\end{equation}
The friction enters the dynamics through its curl and divergence: $\curl\mathbf{F}$
forces $\vort$ in Eq.~\eqref{eq:vort} and $\diver\mathbf{F}$ forces $\divg$ in
Eq.~\eqref{eq:div}.

\codref{chronos\_esm/atmos/dynamics.py}{\code{step\_atmos}: \code{u\_target},
\code{drag\_u}, \code{drag\_v}, \code{curl\_forcing}, \code{div\_forcing}}

\begin{remark}[Why $\tau_f=2$ days, not 6]
The relaxed jet in Eq.~\eqref{eq:utarget} has meridional shear that is itself
barotropically unstable. At a $6$-day timescale, localized eddies grew faster than
they were damped and $|\uu|$ ran to the clamp by $\sim$ day 14. A $2$-day
relaxation damps deviations from the target faster than the instability grows,
pinning the field near the (stable, realistic) climatological jet. So the timescale
is set by a numerical stability margin, not by an eddy lifetime. This is a
parameterization in the strictest sense --- if you want \emph{emergent} jets and
storm tracks, you need the multi-level core (Chapter~\ref{chap:ch18_jets_stormtracks}).
\end{remark}

\section{Physics: precipitation, radiation, CO\texorpdfstring{$_2$}{2}}

The column physics lives in \code{chronos\_esm/atmos/physics.py} and is written to
be differentiable end-to-end (smooth or piecewise-linear triggers, no hard
branches that break the gradient).

\paragraph{Saturation and precipitation.}
Saturation specific humidity follows Clausius--Clapeyron with the Tetens form,
\begin{equation}
  e_s(T) = 611\,\exp\!\Big(\frac{17.27\,(T-273.15)}{(T-273.15)+237.3}\Big),
  \qquad
  q_s = 0.622\,\frac{e_s}{p}.
\end{equation}
Precipitation uses a relative-humidity threshold $q_c = q_c^{\mathrm{ref}}\,q_s$ and
a smoothed convective trigger. The intended form is a softplus,
$P=\softplus\!\big((q-q_c)/\epsilon\big)\,\epsilon/\tau_{\mathrm{conv}}$, but the
code as shipped uses a rectified-linear (ReLU) trigger,
\begin{equation}
  P = \mathrm{ReLU}\!\Big(\frac{q-q_c}{\epsilon}\Big)\,
      \frac{\epsilon}{\tau_{\mathrm{conv}}},
  \qquad
  Q_{\mathrm{lat}} = \frac{L_v}{c_p}\,P,
  \label{eq:precip}
\end{equation}
with $\tau_{\mathrm{conv}}=1800\ \mathrm{s}$, $L_v=2.5\times10^6\
\mathrm{J\,kg^{-1}}$, $c_p=1004\ \mathrm{J\,kg^{-1}\,K^{-1}}$, and
$q_c^{\mathrm{ref}}=0.4915$, $\epsilon=4.71\times10^{-2}$ both tuned (the
docstring's stated target was an idealized ITCZ). The smoothing parameter $\epsilon$
exists precisely so the trigger has a well-defined derivative for the
differentiable-calibration applications of
Chapter~\ref{chap:ch15_differentiable_applications}.

\codref{chronos\_esm/atmos/physics.py}{\code{compute\_saturation\_humidity},
\code{compute\_precipitation}}

\paragraph{Radiation and CO$_2$ forcing.}
Longwave physics is a Newtonian (Newtonian-cooling) relaxation toward a radiative
equilibrium profile $T_{\mathrm{eq}}$, plus a column-distributed CO$_2$ forcing:
\begin{equation}
  Q_{\mathrm{rad}} = -\frac{T - T_{\mathrm{eq}}}{\tau_{\mathrm{rad}}}
                     + \frac{F_{\mathrm{CO_2}}}{\mathcal{C}},
  \qquad
  F_{\mathrm{CO_2}} = \alpha_{\mathrm{CO_2}}\,\ln\!\frac{C}{C_0}.
  \label{eq:rad}
\end{equation}
The CO$_2$ term is the standard logarithmic radiative-forcing law
\citep{myhre1998} with $\alpha_{\mathrm{CO_2}}=5.35\ \mathrm{W\,m^{-2}}$ and
pre-industrial reference $C_0=280$ ppm, divided by a column heat capacity
$\mathcal{C}=10^7\ \mathrm{J\,m^{-2}\,K^{-1}}$ to convert flux to a heating rate.
The radiative-equilibrium target $T_{\mathrm{eq}}$ is driven by the (seasonal)
downward shortwave flux when supplied, $T_{\mathrm{eq}}=245+0.28\,S_\downarrow$
(clipped to $[180,350]$ K), or otherwise by an annual-mean profile
$T_{\mathrm{eq}}=315-60\sin^2\!\lat$ scaled as $(S_0/1361)^{1/4}$.

\codref{chronos\_esm/atmos/physics.py}{\code{compute\_radiative\_forcing},
\code{compute\_solar\_insolation}, \code{compute\_albedo}}

\begin{remark}[An aggressive cooling timescale]
The implemented $\tau_{\mathrm{rad}}=1$ day is far shorter than the $\sim 20$-day
radiative damping the module constant \code{RAD\_COOL\_RATE} advertises. The
comment is explicit about why: the single-level dynamics mix heat \emph{faster}
than a realistic radiative relaxation could maintain the equator--pole gradient, so
the cooling rate is cranked up to force the gradient by brute relaxation. This is a
symptom of the missing vertical structure, not a free physical choice --- and it is
one more reason this model's tropical SST and ITCZ are biased (cold tongue, weak
equatorial rain).
\end{remark}

\paragraph{Surface fluxes.}
Sensible and latent heat use bulk aerodynamic formulae \citep{largeyeager2004},
$\mathrm{SH}=\rho_a c_p C_h\,|\uvec|\,(T_s-T_a)$ and
$\mathrm{LH}=\rho_a L_v C_e\,|\uvec|\,(q_s-q_a)\,\beta$, with a gustiness floor
$|\uvec|\ge 1\ \mathrm{m\,s^{-1}}$ and a cap $|\uvec|\le 25\ \mathrm{m\,s^{-1}}$ to
forestall a wind-evaporation-feedback (WISHE) runaway, and an evaporation
efficiency $\beta$ ($1$ over ocean, $<1$ over land).
\codref{chronos\_esm/atmos/physics.py}{\code{compute\_surface\_fluxes}}

\section{Conservation fixers, clamps, and limiters}

Forward Euler conserves neither energy nor enstrophy, so over century-scale runs
tiny per-step errors compound. The code applies a \emph{gentle} energy fixer that
nudges temperature to offset the spurious change in total energy
$E=\int(\tfrac12|\uvec|^2+c_p T)\cos\lat\,\dd A$,
$T \leftarrow T - \alpha\,\delta T$ with $\alpha=0.01$ (reduced from $0.1$ once the
polar instability was removed). An enstrophy fixer exists analogously; jointly
conserving energy and enstrophy is the design goal of the classic
energy--enstrophy-conserving discretizations \citep{arakawalamb1981}.
\codref{chronos\_esm/atmos/dynamics.py}{\code{apply\_energy\_fixer},
\code{apply\_enstrophy\_fixer}}

For robustness the integrator also imposes physics-stability clamps each step:
$\spechum\in[0,0.05]$, $T\in[150,350]$ K, $|\vort|,|\divg|\le 5\times10^{-4}\
\mathrm{s^{-1}}$, $\lnps\in[9.2,11.7]$ (i.e.\ $p_s\in[100,1200]$ hPa), and
$C\in[0,10^6]$ ppm. Moisture and temperature \emph{tendencies} are also throttled
(e.g.\ $|Q_T|\le 5\times10^{-3}\ \mathrm{K\,s^{-1}}$) to damp rain/heating shocks.
These are safety rails: in a healthy run none should be active, and a clamp that
\emph{is} hit is a diagnostic that something upstream is wrong.

\section{Honest limitations}

The single-level model reproduces a respectable zonal-mean climate --- after the v3
overhaul, $\sigma(p_s)$ ratio improved from $14.7$ to $2.1$, the $p_s$ bias from
$-71.8$ to $+2.9$ hPa, and surface-wind correlation against ERA5 from $-0.08$ to
$0.72$. But two limitations are structural, not tuning issues:
\begin{enumerate}[leftmargin=2em]
  \item \textbf{No baroclinic eddies, hence no synoptic systems.} A barotropic
    layer cannot extract available potential energy from the temperature gradient.
    The mid-latitude $p_s$ and meridional-wind pattern correlations are
    $\sim 0$; the surface westerlies are \emph{imposed}
    (Eq.~\eqref{eq:utarget}), not earned. Storm tracks require the multi-level core
    (Chapter~\ref{chap:ch18_jets_stormtracks}).
  \item \textbf{A weak ITCZ.} With one level there is no Hadley overturning to
    converge moisture at the equator; precipitation is too dry and rains in the
    subtropics rather than at the equator. The aggressive radiative cooling and
    cold tropical SST compound this (Chapter~\ref{chap:ch19_tropics_monsoon}).
\end{enumerate}
Both are the explicit motivation for the multi-level dynamical core of
Chapter~\ref{chap:ch09_atmosphere_multilevel}. The single-level model remains the
production atmosphere for coupled runs because it is cheap, stable, and
differentiable; it is best read as a deliberately minimal baseline against which the
multi-level core's gains are measured.

\section*{Exercises}

\begin{exercise}
Using \code{step\_atmos}, initialize the atmosphere from \code{init\_atmos\_state},
set the momentum-forcing timescale $\tau_f$ in Eq.~\eqref{eq:drag} to
$\infty$ (i.e.\ remove the relaxation by zeroing \code{drag\_u}, \code{drag\_v}),
and integrate for 30 days from a $10\cos\lat\ \mathrm{m\,s^{-1}}$ jet. Plot the
zonal-mean $\uu$ each 5 days. Confirm the prediction of Section~\ref{sec:windrelax}:
the barotropic layer spins down toward rest. Roughly what $e$-folding time do you
measure, and how does it compare with $\tau_f=2$ days?
\end{exercise}

\begin{exercise}
Reproduce the polar blow-up of Section~\ref{sec:polardx}. Temporarily replace
$\Delta x_{\mathrm{dyn}}$ in Eq.~\eqref{eq:dxfloor} by the bare
$2\pi a\cos\lat/\mathrm{nlon}$ (no floor), integrate from rest with flat
topography, and record $\sigma(p_s)$ at the pole rows versus the floored run. How
many steps until $p_s$ leaves the $[100,1200]$ hPa clamp? This is a hands-on
demonstration of why a lat--lon dynamical core needs a polar treatment.
\end{exercise}

\begin{exercise}
The CO$_2$ forcing in Eq.~\eqref{eq:rad} is logarithmic
\citep{myhre1998}. Holding all dynamics fixed, drive
\code{compute\_radiative\_forcing} with $C\in\{280,560,1120\}$ ppm and verify that
each doubling adds a constant $\alpha_{\mathrm{CO_2}}\ln 2\approx 3.7\
\mathrm{W\,m^{-2}}$ to $F_{\mathrm{CO_2}}$. Then use \jax's \code{jax.grad} to
compute $\partial T_{\mathrm{eq,column}}/\partial C$ at $C=400$ ppm --- this is the
kind of exact sensitivity that motivates the differentiable applications of
Chapter~\ref{chap:ch15_differentiable_applications}.
\end{exercise}

\begin{exercise}
The precipitation trigger Eq.~\eqref{eq:precip} ships as a ReLU even though the
docstring describes a softplus. (a) Show analytically that
$\softplus(x)\to\mathrm{ReLU}(x)$ as the softplus sharpness increases, and that the
two differ by at most $\ln 2$ at $x=0$. (b) Swap the ReLU for a true softplus in
\code{compute\_precipitation}, run a short integration, and report the change in
global-mean precipitation and in the smoothness of $\partial P/\partial q$ near the
threshold. Which form is friendlier to gradient-based calibration, and why might
the ReLU still have been chosen?
\end{exercise}
