\chapter{Teaching with Chronos-ESM}\label{chap:ch21_teaching}

\providecommand{\labweek}[1]{\textbf{Week #1}}

Most graduate climate courses teach Earth System Models the way one teaches an opaque
instrument: students read the governing equations on a chalkboard, then run a community model
(CESM, MOM6, an EC-Earth configuration) as a black box on a remote queue, waiting hours or days
for output they did not generate and cannot fully read. The pedagogy and the software pull in
opposite directions --- the equations are transparent but inert, the model is alive but
inscrutable. \model{} was not built to be a teaching model, but several of its design choices
make it an unusually good one. This chapter argues why, lays out a 12--14 week graduate course
that uses the model as a weekly laboratory, gives concrete lab assignments drawn from the
\emph{real} experiment scripts in the repository, and closes with notes for instructors.

A standing caveat frames everything below. \model{} is research software --- a research preview,
not a production climate model (Chapter~\ref{chap:ch01_introduction}). Some components are
validated against observations (ocean T/S structure, the single-level surface winds, the
multi-level ITCZ); others are explicit simplifications or carry documented biases (the
single-level atmosphere has no synoptic eddies; the interim AMOC closure is a box-model
parameterization, not prognostic momentum). This is a feature for teaching, not a bug: a model
whose limitations are written down is a model whose limitations a student can \emph{discover},
which is the most durable lesson a climate course can deliver.

\section{Why \model{} teaches well}

\subsection{Minutes-to-hours runtime closes the experiment loop within one session}

The single most important pedagogical property of \model{} is speed. On a single GPU the coupled
model integrates roughly $280$ simulated years per day (Chapter~\ref{chap:appD_running}), so a
multi-decadal control run finishes in tens of minutes and a short forced experiment in single
minutes. Even on a laptop CPU the differentiable kernels are fast enough for the short probes the
labs below use. The consequence is that a student can \emph{conceive, run, plot, and interpret} a
numerical experiment inside one three-hour lab. The hypothesis-to-result loop --- the actual
engine of scientific learning --- closes before the student leaves the room, rather than across a
week-long queue wait during which the original question is forgotten. This is the difference
between an experiment a student \emph{owns} and a figure a student \emph{receives}.

\subsection{Full source legibility}

\model{} is a few tens of thousands of lines of readable \jax{}/Python, not millions of lines of
Fortran accreted over three decades. A student can open \code{chronos\_esm/ocean/veros\_driver.py}
and read, in one sitting, the actual tracer advection their experiment exercises; they can open
\code{chronos\_esm/ocean/overturning.py} and see the exact line where the salt-advection feedback
gain enters (Chapter~\ref{chap:ch08_ocean_overturning}). The equation on the lecture slide and the
line of code that integrates it are separated by minutes of reading, not by an institutional
firewall. This collapses the chalkboard/black-box gap described above: the model \emph{is} the
lecture notes, executable. Throughout the book the \emph{Code:} margin pointers exist precisely so
a reader can jump from a result to its source.

\subsection{Differentiability makes sensitivity and inverse methods first-class}

\model{} is end-to-end differentiable (Chapters~\ref{chap:ch03_differentiable},
\ref{chap:ch15_differentiable_applications}): \jax{}'s reverse-mode autodiff
\citep{jax2018} returns the exact gradient of any scalar diagnostic with respect to any input ---
a parameter, a boundary condition, an initial state. In a conventional course, ``sensitivity'' is
taught as a finite-difference afterthought (run the model twice, subtract, divide). Here a student
can compute $\partial(\text{AMOC})/\partial(\text{subpolar salinity})$ as a single backward pass
and cross-check it against a central finite difference in the same script
(\code{experiments/verify\_gradient.py}, \code{experiments/p0\_amoc\_gradient\_baseline.py}). This
turns abstract topics --- adjoint sensitivity, 4D-Var data assimilation, gradient-based parameter
calibration, the envelope theorem at a bifurcation --- into things a student can \emph{run}. Few
teaching models can do this at all; fewer can do it in a script short enough to read in a lab.

\section{A 12--14 week graduate course}

The course below maps one lecture block per week to the book's chapters and pairs each with a
computer lab that uses the model. It is sized for a one-semester graduate course (12 core weeks
plus 2 optional project weeks) and assumes a parallel track of problem sets on the analytic
material. The arc is deliberate: build the ocean first (it is the validated core), add the
atmosphere, couple, then spend the back half on the three things differentiability and speed make
uniquely teachable --- forcing response, tipping, and inverse methods.

\begin{description}[leftmargin=0pt,style=nextline]
  \item[\labweek{1} --- Orientation and the differentiable paradigm.]
    Lecture: Chapters~\ref{chap:ch01_introduction}, \ref{chap:ch02_preliminaries},
    \ref{chap:ch03_differentiable}. What an ESM is; the model hierarchy; why ``differentiable''
    changes what you can ask. Lab: install, run a short control
    (\code{run\_century\_physics.py --years 5}), open three source files and trace one tracer
    through the code. Deliverable: a plot of global-mean SST settling, with a paragraph on what
    the student read in the source.
  \item[\labweek{2} --- Ocean governing equations.]
    Lecture: Chapter~\ref{chap:ch04_ocean_equations}. Primitive equations, hydrostatic and
    Boussinesq approximations, the equation of state. Lab~A (below): perturb the EOS and watch the
    circulation respond.
  \item[\labweek{3} --- Ocean mixing and parameterization.]
    Lecture: Chapter~\ref{chap:ch05_ocean_mixing}. Vertical mixing, isopycnal diffusion, the
    Gent--McWilliams eddy parameterization \citep{gentmcwilliams1990}. Lab: vary \code{kappa\_gm}
    and \code{kappa\_h} in a short control; compare the resulting thermocline depth.
  \item[\labweek{4} --- Ocean numerics.]
    Lecture: Chapter~\ref{chap:ch06_ocean_numerics}. The C-grid, flux-form advection, the
    z-level staircase bathymetry, stability and the CFL condition \citep{griffies2004}. Lab:
    halve the time step and the grid masking; observe stability limits and conservation.
  \item[\labweek{5} --- The prognostic ocean core.]
    Lecture: Chapter~\ref{chap:ch07_ocean_prognostic_core}. Diagnostic versus prognostic
    momentum; why the current ocean's velocities are reconstructed and what that costs
    (the $\partial(\text{AMOC})/\partial\dens\approx0$ blocker). This is the week to be honest
    about a research model's frontier. Lab: run \code{p0\_amoc\_gradient\_baseline.py} and
    measure the blocker directly.
  \item[\labweek{6} --- Overturning and the AMOC.]
    Lecture: Chapter~\ref{chap:ch08_ocean_overturning}. The overturning streamfunction; the
    Stommel two-box model \citep{stommel1961}; the interim density-driven closure. Lab~C (below):
    hose the AMOC.
  \item[\labweek{7} --- The atmosphere I: multi-level dynamics.]
    Lecture: Chapter~\ref{chap:ch09_atmosphere_multilevel}. Spectral primitive equations,
    sigma coordinates, baroclinic instability, the \dino{} dycore behind NeuralGCM
    \citep{kochkov2024neuralgcm}; the Held--Suarez idealized benchmark \citep{heldsuarez1994}.
    Lab: run \code{dino\_held\_suarez.py}; recover the twin midlatitude jet from rest.
  \item[\labweek{8} --- The atmosphere II: the single level and its limits.]
    Lecture: Chapter~\ref{chap:ch10_atmosphere_singlelevel}. The barotropic single level, the
    eddy-momentum-flux relaxation, why a single level has no synoptic systems. Lab: score the
    atmosphere against ERA5 with \code{validate\_control.py --demo}; read the skill table
    honestly (where it wins, where it fails).
  \item[\labweek{9} --- Sea ice, land, and the coupler.]
    Lecture: Chapters~\ref{chap:ch11_seaice}, \ref{chap:ch12_land},
    \ref{chap:ch13_coupler_fluxes}. Thermodynamic ice, the bucket land model, bulk-formula
    surface fluxes, flux correction. Lab: run a short coupled \dino{}$\leftrightarrow$ocean
    control (\code{run\_dino\_coupled.py --days 30}) and inspect the surface energy budget.
  \item[\labweek{10} --- Forcing and response.]
    Lecture: Chapter~\ref{chap:ch14_forcing_response}. Radiative forcing of CO$_2$
    \citep{myhre1998}, transient versus equilibrium response, feedbacks, why \model{}'s warming is
    a forcing-vs-warming \emph{proxy} not an ECS. Lab~D (below): double CO$_2$.
  \item[\labweek{11} --- Differentiable applications.]
    Lecture: Chapter~\ref{chap:ch15_differentiable_applications}. Adjoint sensitivity,
    gradient-based calibration, 4D-Var, calibrating a bifurcation by AD fold-continuation.
    Lab~E (below): take a gradient and an inverse.
  \item[\labweek{12} --- Evaluating a model: climatology and CMIP.]
    Lecture: Chapters~\ref{chap:ch16_climatology}, \ref{chap:ch20_cmip7}. Skill metrics, Taylor
    diagrams, model intercomparison \citep{eyring2016cmip6}. Lab~B (below): build and read a
    validation scorecard.
  \item[\labweek{13--14} --- Selected phenomena / project work.]
    Lecture (pick by interest): Chapters~\ref{chap:ch17_amoc_tipping} (tipping),
    \ref{chap:ch18_jets_stormtracks} (jets and storm tracks),
    \ref{chap:ch19_tropics_monsoon} (tropics). Lab: supervised work on the term project
    (Section~\ref{sec:teach-project}). Reference appendices throughout:
    \ref{chap:appA_notation} (notation), \ref{chap:appB_parameters} (parameters),
    \ref{chap:appC_codemap} (code map), \ref{chap:appD_running} (running).
\end{description}

\section{Lab assignments}

The six labs below are drawn directly from runnable scripts in \code{experiments/}. Each gives a
learning goal, the command to run, what to plot, and discussion questions. All are sized to finish
within one lab session. Commands assume the repository root and an activated environment
(Chapter~\ref{chap:appD_running}); paths and flags match the scripts as shipped.

\subsection{Lab A --- Run a control and perturb the equation of state}

\textbf{Learning goal.} Connect the equation of state to the large-scale circulation: density
sets pressure gradients, which drive flow. Establish what a ``control'' run is and how to read a
spin-up.

\textbf{Run.}
\begin{quote}\small
\code{python experiments/run\_century\_physics.py --years 10 --ocean-ic woa}
\end{quote}
Then re-run with a perturbed thermal-expansion or haline-contraction coefficient in the EOS
(\code{chronos\_esm/ocean/}; a one-line edit the source-legibility property makes easy), saving to
a second output directory.
\codref{experiments/run\_century\_physics.py}{WOA-seeded control / spin-up}

\textbf{Plot.} Global-mean SST and SSS time series for control versus perturbed; a surface-current
or streamfunction map for each; the difference map.

\textbf{Discussion.} Which way did the gyres shift when you made seawater more (or less) sensitive
to temperature? Did the perturbation change the \emph{mean} state or the \emph{spatial pattern}
more? The control still relaxes salinity in a high-latitude sponge --- how does that constrain
what your perturbation can do?

\subsection{Lab B --- Score the model against observations}

\textbf{Learning goal.} Learn what ``validation'' means quantitatively: bias, RMSE, pattern
correlation, standard-deviation ratio, and the Taylor framework
(Chapter~\ref{chap:ch16_climatology}). Learn to read a model's skill \emph{honestly}.

\textbf{Run.}
\begin{quote}\small
\code{python experiments/validate\_control.py --demo 200 --era5}
\end{quote}
(or, against saved checkpoints, \code{--states "outputs/century\_physics/year\_*.nc" --era5}).
\codref{experiments/validate\_control.py}{score a run against WOA18 + ERA5}

\textbf{Plot.} The auto-generated scorecard table, the bias maps, and the zonal-mean comparisons
the script writes.

\textbf{Discussion.} The single-level atmosphere scores surface zonal-wind correlation near $0.7$
but mean-sea-level-pressure pattern correlation near $0$ --- why does it get winds right and
pressure wrong? (Hint: eddies.) Which fields are \emph{predicted} and which are \emph{imposed}
(in the flux-corrected coupled control, SST correlation is $1.00$ by construction)? Pick the
field you trust least and defend why.

\subsection{Lab C --- Hose the AMOC and find a tipping point}

\textbf{Learning goal.} Reproduce a saddle-node bifurcation with hysteresis from the
salt-advection feedback \citep{stommel1961}; understand why a rate-dependent sweep is ambiguous
and a fixed-forcing on/off test is rigorous (Chapter~\ref{chap:ch17_amoc_tipping}).

\textbf{Run.} A short self-contained quasi-static sweep:
\begin{quote}\small
\code{python experiments/run\_amoc\_hosing.py --spinup-years 20 --fmax 0.8 --nsteps 6}\\
\code{\ \ --hold-years 15 --haline-gain 6 --k-vel 6e-5 --contrast-depth 300}
\end{quote}
\codref{experiments/run\_amoc\_hosing.py}{quasi-static subpolar freshwater hosing sweep}

\textbf{Plot.} AMOC at $26.5^\circ$N versus hosing $F$ for the up-leg and down-leg on one axis ---
the hysteresis loop. The calibrated configuration opens a window $\approx[0.38,\,0.75]\Sv$
centered near $0.6\Sv$.

\textbf{Discussion.} Why does the down-leg sit below the up-leg over a finite $F$ window, and what
physical feedback makes the ``off'' state self-sustaining? The branches are noisy
($\pm\sim10\Sv$) --- is that numerical noise or a genuine relaxation oscillation, and how would
you tell? Why is the interim closure a box-model parameterization rather than prognostic momentum,
and what would a clean bifurcation require (Chapter~\ref{chap:ch07_ocean_prognostic_core})?

\subsection{Lab D --- Double CO$_2$ and measure the warming}

\textbf{Learning goal.} Separate radiative \emph{forcing} from climate \emph{response}
\citep{myhre1998}; understand what a transient forcing-vs-warming slope is and why it is
\emph{not} equilibrium climate sensitivity in this model.

\textbf{Run.} From a converged control checkpoint, two branches (control ppm and $2\times$):
\begin{quote}\small
\code{python experiments/run\_dino\_co2.py --ckpt <control\_ckpt> --co2 560 --years 50}
\end{quote}
\codref{experiments/run\_dino\_co2.py}{abrupt-2$\times$CO$_2$ on the freed q-flux model}

\textbf{Plot.} The two SST time series and their difference $\Delta\text{SST}(t)=$ forced $-$
baseline (drift cancels in the difference).

\textbf{Discussion.} The script reports $\Delta\text{SST}\approx+1.6$~K, about
$0.43$~K per W\,m$^{-2}$. Why is this a \emph{proxy} and not ECS? (No top-of-atmosphere energy
closure, no atmospheric radiative-feedback amplification, a cold-tropics base state.) Which
feedbacks present in a full GCM are absent here, and which way would each move the number? Why is
taking the \emph{difference} of two branches more trustworthy than either branch alone?

\subsection{Lab E --- Take a gradient and run an inverse}

\textbf{Learning goal.} Use differentiability directly: compute an exact sensitivity by
reverse-mode autodiff \citep{jax2018}, cross-check it against finite differences, and use a
gradient to \emph{calibrate} a target (Chapter~\ref{chap:ch15_differentiable_applications}).

\textbf{Run.} The sensitivity baseline, then a calibration:
\begin{quote}\small
\code{python experiments/p0\_amoc\_gradient\_baseline.py}\\
\code{python experiments/calibrate\_amoc\_fold.py --target 0.6 --kvel-rel 0.6}
\end{quote}
\codref{experiments/p0\_amoc\_gradient\_baseline.py}{$\partial(\text{AMOC})/\partial(\text{density})$ by grad + finite difference}
\codref{experiments/calibrate\_amoc\_fold.py}{AD fold-continuation calibration of $F_{\mathrm{crit}}$}

\textbf{Plot.} The grad value next to the finite-difference value (agreement check); the
calibrated $(\code{haline\_gain},\code{k\_vel})$ that places the fold at the target $F_{\mathrm{crit}}$.

\textbf{Discussion.} The baseline finds $\partial(\text{AMOC})/\partial(\text{density})\approx0$
on the diagnostic-velocity ocean --- what does a \emph{zero} sensitivity tell you about the
model's physics, and why is recording a zero a legitimate scientific result? Why can a tipping
threshold \emph{not} be tuned by backpropagating through the hosing sweep itself (the sensitivity
diverges \emph{at} the fold), and how does fold-continuation via the envelope theorem get an exact
gradient anyway? Rank \code{k\_vel} versus \code{haline\_gain} as levers from the gradient.

\subsection{Lab F --- Recover a jet from rest (idealized dynamics)}

\textbf{Learning goal.} See baroclinic instability and the eddy-driven jet emerge from a
controlled idealized forcing, decoupled from observational boundary conditions
\citep{heldsuarez1994} (Chapters~\ref{chap:ch09_atmosphere_multilevel},
\ref{chap:ch18_jets_stormtracks}).

\textbf{Run.}
\begin{quote}\small
\code{python experiments/dino\_held\_suarez.py}
\end{quote}
\codref{experiments/dino\_held\_suarez.py}{\dino{} dry dycore with Held--Suarez forcing}

\textbf{Plot.} The zonal-mean zonal wind: twin upper-tropospheric midlatitude westerlies
($\sim22$~m/s) over surface trades.

\textbf{Discussion.} Where does the eddy momentum flux that builds the surface westerlies come
from, and why can the single-level model (Lab~B) not produce it? Why does the textbook jet emerge
here from an \emph{isothermal} start but the realistic coupled run needs a symmetry-breaking seed?
What does Held--Suarez deliberately leave out, and why is that the point?

\section{An assessment / term-project idea}
\label{sec:teach-project}

\textbf{Project: a small intercomparison of one tunable knob.} Each student (or pair) picks one
model parameter --- the GM coefficient \code{kappa\_gm}, the salinity-sponge timescale, the
\code{haline\_gain}, the diffusion $\tau$ of the multi-level atmosphere --- and conducts a
miniature, self-contained intercomparison in the spirit of CMIP \citep{eyring2016cmip6}. The
deliverable is a short paper with: (i) a physical hypothesis for how the knob should change the
climate; (ii) a 3--5 member ensemble of short runs spanning the knob; (iii) a validation scorecard
(Lab~B) for each member; (iv) where differentiability applies, the \emph{gradient} of one target
diagnostic with respect to the knob, cross-checked against the ensemble's finite difference; and
(v) an honest discussion of which observed feature the knob improved, which it degraded, and what
the model's documented limitations mean for the result. The runtime budget makes a real ensemble
feasible inside a term, and the validation framework supplies a common, reproducible scoring
standard so the class can pool results into one Taylor diagram.

A lighter alternative for a shorter course: reproduce one published figure from the book
(the AMOC hysteresis loop, the Held--Suarez jet, the $2\times$CO$_2$ warming curve) from scratch,
and write a one-page critique of what the figure does and does not establish.

\section{Notes and exercises for instructors}

A teaching model is only as good as the honesty with which it is taught. \model{}'s greatest
classroom value is that its failures are documented and reproducible: a student who measures the
$\partial(\text{AMOC})/\partial\dens\approx0$ blocker, or reads the near-zero MSLP pattern
correlation, has learned more about how models work than one handed a polished CESM figure. Design
the labs so the model is allowed to be wrong, and grade the interpretation, not the agreement.

\begin{description}[leftmargin=0pt,style=nextline]
  \item[Instructor Exercise 1 --- Calibrate the difficulty of Lab~C.]
    Before assigning the hosing lab, run the sweep yourself at two or three \code{haline\_gain}
    values (e.g.\ $4$, $6$, $8$) and time each on your teaching hardware. Identify a configuration
    that (a) opens a clearly visible hysteresis loop and (b) finishes inside your lab session.
    Write a half-page rubric distinguishing a student who found a \emph{real} loop from one who
    mistook the $\pm10\Sv$ branch noise for one.
  \item[Instructor Exercise 2 --- Build an answer key for the honest-validation discussion.]
    Run Lab~B yourself and, for each of the seven scored fields, write one sentence stating
    whether the model's skill is \emph{earned}, \emph{imposed}, or \emph{absent}, with the
    physical reason. (Worked starting points: SST in the flux-corrected control is imposed; the
    ITCZ precipitation in the multi-level atmosphere is earned; the MSLP pattern is absent for
    lack of synoptic eddies.) Use it to grade interpretation rather than memorized numbers, which
    drift as the model is refined.
  \item[Instructor Exercise 3 --- Map your syllabus onto the model's frontier.]
    The model is actively developed; some labs above exercise validated components and some
    exercise the research frontier (the prognostic ocean core, the multi-level atmosphere's
    surface winds). Before the term, classify each lab as \emph{settled} or \emph{frontier} for
    your offering, and decide explicitly which frontier results you will present as
    open questions rather than as answers. Document one place where the model has changed since
    this chapter was written, and adjust the affected lab.
\end{description}

The recurring meta-lesson across all three is the same one the labs teach the students: an Earth
System Model is an argument, not an oracle. Teaching with a model whose source you can read, whose
runs finish before the bell, and whose gradients you can take, lets a student examine that
argument from the inside --- which is the whole point.
